\chapter{Introduction}\label{chapter:intro}%
In this thesis we propose a novel shift-and-invert method to compute
an eigenvector and corresponding eigenvalue of a real symmetric
matrix. The method is a modification of the \emph{Rayleigh Quotient
  Iteration} (RQI) designed to overcome some of the drawbacks of
RQI. It is well-known that RQI is often unpredictable in the sense
that even when executed with accurate eigenvector approximations it
does not necessarily yield the correct eigenpair. This problem becomes
worse in cases where the distance between the wanted eigenvalue and
its neighbours is very small. In our new method we perturb the
original problem in such a way that the spacing between the target
eigenvalue and the other eigenvalues is increased and apply RQI to
this perturbed problem. Obviously, this approach alone would lead to
incorrect results. Therefore, the perturbation is successively
decreased until we finally arrive at the original problem and we hope
that by that time the current eigenvalue approximation has become
sufficiently close to the correct result so that RQI succeeds. For
this approach to work, a sufficiently good eigenvector approximation
is needed and our numerical experiments suggest that ``sufficiently
good'' means that the angle between the initial approximation and the
target eigenvector is below $45^\circ$---independent of the spacing of
the eigenvalues. Although the method converges not as fast as RQI,
often merely three or four additional iterations are required compared
to RQI.

The thesis is structured as follows. In this chapter we collect some
basic definitions and results from numerical linear algebra. We also
review some simple iterative methods for eigenvalue problems. Since
our new algorithm is strongly related to RQI, the second chapter is
devoted to a detailed study of the Rayleigh Quotient and RQI. Among
others, we give a proof of the property that makes RQI advantageous
over other methods, viz., the local cubic convergence. In the third
chapter we introduce our novel method that we call \emph{Complex
  Rayleigh Quotient Iteration}. After motivating the method, different
numerical experiments are carried out for various test problems to
better understand its behaviour and to illustrate the good properties
mentioned above.

The aim of this thesis is not to carry out a mathematical analysis of
the method but rather to discuss various numerical examples that
illustrate different behavioural aspects of the algorithm. We leave a
detailed analysis of the method based on our observations for future
work.

% \input{Chapters/01/objectives}%
\input{Chapters/01/SymmetricEigvalProblem}%
\input{Chapters/01/IterativeMethods}%

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../main"
%%% End:
