\chapter{Classic Rayleigh Quotient Iteration}\label{chapter:rqi}
In this chapter we introduce the Rayleigh Quotient Iteration (or RQI,
for short) and discuss some important results. Among others, we give a
proof of the most appealing property of RQI, namely the (local) cubic
convergence.

As mentioned above, we want to find the eigenvalues and eigenvectors
of a real symmetric matrix $\mat{A}$. We make the following additional
assumptions:
\begin{enumerate}[label=(A\arabic*),leftmargin=38pt]
\item The matrix $\mat{A}$ is large.
\item A good approximation of the wanted eigenvector is available.
\item The wanted eigenvalue lies in the interior of the spectrum.
\item The eigenvalues around the wanted eigenvalue are closely spaced.
\end{enumerate}

The first assumption prohibits us from using an algorithm that
computes the full eigendecomposition. Due to the second assumption we
would like a method that makes use of the approximation in the sense
that it is faster the more accurate the approximation is. Assumption
(A3) seems to pose no real constraint at first sight. However, many of
the different eigenvalue algorithms are only capable of finding a few
eigenvalues at either end of the spectrum or they are at least
considerably faster/more likely to succeed in these cases.

It will turn out that RQI is a suitable choice if we omit the last
assumption. Although a different linear system has to be solved at
each step, RQI often converges within five steps when stared with a
sufficiently good eigenvector approximation. The number of steps does
not necessarily increase much when the eigenvalues are closely
spaced. However, RQI usually fails to compute the right eigenpair in
this cases and it is generally impossible to predict which eigenpair
the method will converge to. The main goal of the next chapter is to
design a method that overcomes this problem.

% 
\input{Chapters/02/TheRayleighQuotient}%
\input{Chapters/02/History}%
\input{Chapters/02/Convergence}%

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../main"
%%% End:
